---
title: "Reproducible Research- Assignment 1"
author: "ME"
date: "February 20, 2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
```

# Assignment 1
##Step 1: Loading the data

The following code loads the pre-downloaded dataset into R into a object called "dataset". 
```{r download_data}
        dataset<-read_csv("activity.csv")

```


##Step 2: What is the mean total number of steps taken per day?

1.Calculate the total number of steps taken per day?  
        -using dplyr functions to summarize data and grouped data is stored in group1 variable  
```{r stepcount}
        group1<-dataset%>%group_by(date)%>%summarize(totalsteps=sum(steps))
        head(group1)
        
```

2.Make a Histogram of the total number of steps taken each day.  
        -Histogram is plot using ggplot2 package   
```{r Histogram plot}
        plot1<-ggplot(group1,aes(totalsteps))+geom_histogram(binwidth=1000)
        print(plot1)
        #ggsave("plot1.png")
```

3.Calculate and report the mean and median of the total number of steps taken each day  
        -Mean is calculated as follows with NA days removed:
```{r mean}
        mean(group1$totalsteps,na.rm=T)
```
-Median is calculates as follows with NA days removed:
```{r median}
        median(group1$totalsteps,na.rm=T)
```

##Step3: What is the average daily activity pattern?  

1. Make a time series plot of the 5-minute interval and the average number of steps taken averaged across all days.  
        -using dplyr again to group by interval and summarize to get average
```{r line chart}
        group2<-dataset%>%group_by(interval)%>%summarise(averagesteps=mean(steps,na.rm=T))

        plot2<-ggplot(group2,aes(interval,averagesteps))+geom_line()
        print(plot2)
        #ggsave("plot2.png")
```  

2.Which 5 minute interval on average contains the maximum number of steps?  
        -using subsetting and "which" function:
```{r maxsteps}
        group2[which(group2$averagesteps==max(group2$averagesteps)),]
```   

##Step4: Inputting missing values...  

1. Calculate and report total number of missing values in dataset.  
        -create variable that filters to NAs (which are only found in steps variable)  
        -sum column of NAs
```{r countNAs}
        NAs<-which(is.na(dataset))
        group3<-dataset[NAs,]
        sum(is.na(group3$steps))
```
2. Devise a strategy for filling in missing values inthe dataset  
        -Populating missing values with interval mean  
        
```{r populate NAs}
        #determine interval mean
        group4<-dataset%>%group_by(interval)%>%summarize(avesteps=mean(steps,na.rm=T))
        #create new column on group3 table (dataset filtered to NAs)
        group3<-left_join(group3,group4,by="interval")
        group3<-group3[,2:4] #drops steps column (with all NAs)
        head(group3)
```

3. Create a New Dataset that is equal to the original dataset but with NAs filled in
        -Populate NA indexed rows with replacements
``` {r Cleaned Data}
        
        dataset_clean<-dataset
        dataset_clean[NAs,"steps"]<-group3$avesteps
        head(dataset_clean)

```

4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total steps each day.        
        -Histogram plotted same as in step 2  
```{r Histogram_clean}
        group5<-dataset_clean%>%group_by(date)%>%summarize(totalsteps=sum(steps))

        plot3<-ggplot(group5,aes(totalsteps))+geom_histogram(binwidth=1000)
        print(plot3)
        #ggsave("plot3.png")

```
       
-Mean calculated as follows:
```{r mean_clean}
        mean(group5$totalsteps)
```

-Median calculated as follows:
```{r median_clean}
        median(group5$totalsteps)
```

As the results show, inputting the interval average into the missing observations appears to leave the mean unchanged and the median only marginally shifted.  

##Step5: Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels "weekday" and "weekend"  
        -use wday() function (from lubridate) to add "weekdays" variable defining day  
        -create function to check "weekday" and assign factor, sapply over data  
        -group data by interval and day class, average steps and plot  
        
```{r weekday}
        dataset_clean$weekdays<-wday(dataset_clean$date)
        day_class<-function(x) if (x==1|x==7) {"Weekend"} else {"Weekday"}
        
        dataset_clean$dayclass<-as.factor(sapply(dataset_clean$weekdays,day_class))
        group6<-dataset_clean%>%group_by(dayclass,interval)%>%summarise(Avesteps=mean(steps))
        
        plot4<-ggplot(group6,aes(interval,Avesteps))+geom_line()+facet_wrap(~dayclass,ncol=1)+labs(y="Ave                 Steps",x="Time Interval")
        
        print(plot4)
        #ggsave("plot4.png")

```
        